### 2024-01-23 parts/transducers.tex
avancées

The goal of this section is to provide tools that do not require
\kl{commutative} \kl{$\Nat$-polyregular functions}. To that end, we introduce
the notion of \kl{$k$-residual transducer}, that generalizes the notion of
residual transducer introduced for \kl{$\Rel$-polyregular functions} in
\cite{LOPEZ23b}. We will first introduce the object, provide an algorithm (with
an oracle function) that computes said object, and finally, check that in
simple cases, the aperiodicity of a function can be read on its
\kl{$k$-residual transducer}. The main drawbacks of the generalisation are the
following: one needs to decide whether a \kl{$\Rel$-polyregular function} is in
$\NPoly$, which is only known in the \kl{commutative} case, and the
characterization in terms of counter free also assumes \kl{commutativity}.
However, we believe that the introduction of this object is relevant.

\AP
Let us first introduce the notion of \intro{residual} of a function
$f \colon \Sigma^* \to \Rel$, by defining
$\app{f}{u} \defined w \mapsto f(uw)$.
The collection of \reintro{residuals} of a function $f$
is denoted $\Res(f)$ and is defined as
the set of $\app{f}{u}$ where $u$ ranges over words in $\Sigma^*$.

\AP Let $k \in \Nat$. The residual transducer in the case of $\Rel$ output was
built using an equivalence relation defined as $f \equiv_k g$ if and only if $g
- f \in \ZPoly[k-1]$. However, in the case of $\Nat$ output, the relation
cannot be symmetric, and we are going to introduce a partial ordering $f
\natleq[k] g$ if and only if $g - f \in \NPoly[k]$.

@@

\AP Let us recall that a counter in an automaton is a pair $u,w$ of words
together with an integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0,
u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.

@@

Let us now introduce the notion of residual transducer, based on the
notions of ordering introduced prior to this paragraph.

\begin{definition}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
    is a \intro{k-residual transducer}
    of $f$ 
    when
    \begin{itemize}
        \item $\aTransd$ \kl{computes} $f$;
        \item $Q \subseteq \Sigma^*$ is \kl{downwards closed}
            for $\prefleq$;
        \item For all $u, v \in Q$, and for all $a \in \Sigma$,
            if
            $\delta(v,a) = u$
            then
            $u \prefleq va$,
            $\lambda(v,a) = \app{f}{va} - \app{f}{u}$,
            and $\lambda(v,a) \in \NPoly[k-1]$.
    \end{itemize}
\end{definition}

\begin{fact}
    The residual transducer is unique.
\end{fact}

We introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
given a function $f$. Notice that this algorithm requires the ability to test
if a function belongs to $\NPoly[k]$, which is only known to be feasible for
\kl{commutative} polyregular functions. However, the termination of this
algorithm also proves the existence of a \kl{$k$-residual transducer}.

@@

        \eIf{$\exists v \in Q, v = \max_{\prefleq} \setof{w \in Q}{w \resleq{f}{k} ua}$}{
            $\delta(u, a) \defined v$;

@@

            $O \defined O \cup \setof{ uab }{b \in \Sigma}$;
        }
    }
    \For{$u \in Q$}{
        $F(u) \defined f(u)$;
    }
    \Return{$(Q, \varepsilon, \delta, \lambda, F)$};
    \caption{Computing a $(N,k)$-residual transducer.}
    \label{residual:algo}
\end{algorithm}

\begin{lemma}
    \label{correct-residual:lemma}
    If \cref{residual:algo} terminates on 
    an input $f \in \NPoly$, $k \in \Nat$,
    then it computes a \kl{$k$-residual transducer} of $f$.
\end{lemma}
\begin{proof}
    TODO.
\end{proof}

\begin{fact}
    \label{q-o-prefix-cool:fact}
    Let $f \in \NPoly$ and $k \in \Nat$.
    At each step of the \texttt{while loop}
    of \cref{residual:algo}, the sets
    $Q$ and $O$ are such that
    \begin{enumerate}
        \item $Q \cup O$ is a \kl{downwards closed} subset of 
            $\Sigma^*$ for $\prefleq$;
        \item elements in $O$ are pairwise incomparable
            for $\prefleq$, and are maximal
            for $\prefleq$ inside $Q \cup O$.
    \end{enumerate}
\end{fact}
\begin{proof}
    Let us write $Q_i$ and $O_i$ for the value of the variables
    $Q$ and $O$ at step $i$ of the \texttt{while loop}.
    We prove the desired property by induction on $i$.

@@

\begin{lemma}
    \label{wqo-implies-termination:lemma}
    Let $f \in \NPoly$, and $k \in \Nat$ be such that
    $(\Sigma^*, \resleq{f}{k} \cup \preforth)$ is a \kl{well-quasi-ordering}.
    Then, \cref{residual:algo} terminates on the input $(f,k)$.
\end{lemma}
\begin{proof}
    Assume towards a contradiction that
    \cref{residual:algo} does not terminate.
    Then, the \texttt{else} branch in the \texttt{while loop}
    must be taken infinitely often.
    This means that the set $Q$ of states grows arbitrarily large.

@@

    Let us prove that this infinite branch is a \kl{bad sequence} for the
    ordering $\preforth \cup \resleq{f}{k}$. Because all elements in the
    sequence are pairwise comparable for $\prefleq$, it suffices to prove that
    the sequence is a \kl{bad sequence} for $\resleq{f}{k}$.

    Let $j < p$, and assume by contradiction that $u_j \resleq{f}{k} u_p$. We
    know that $u_j \in Q_j$ and $u_p \in Q_p$. Then, at step $p-1$ of the
    algorithm, $u_j \in Q_{p-1}$, since $u_j \in Q_j \subseteq Q_{p-1}$.
    Because $u_j \prefleq u_p$ and $u_j \resleq{f}{k} u_p$,
    \cref{residual:algo} must take the \texttt{if} branch at step $p-1$. As a
    consequence, $u_p \not\in Q_{p}$, which is absurd.

    We have proven that the infinite branch is a \kl{bad sequence}
    for $\resleq{f}{k}$, but this contradicts the assumption
    that $\seqof{u_j}{j \in \Nat}$ is a \kl{good sequence}.

    Hence, \cref{residual:algo} must terminate.
\end{proof}

\begin{lemma}
    \label{n-poly-k-implies-wqo:lemma}
    Let $k \in \Nat$, 
    let $\aTransd$ be a \kl{$\NPoly[k-1]$-transducer},
    and let $f$ be the function $f$ \kl{computed} by $\aTransd$.
    Then,
    $(\Sigma^*, \resleq{f}{k-1} \cup \preforth)$ is
    \kl{well-quasi-ordering}.
\end{lemma}
\begin{proof}
    Assume towards a contradiction that there exists
    a \kl{bad sequence} $\seqof{u_i}{i \in \Nat}$
    of words in $\Sigma^*$.
    In particular, for all $i < j$, 
    $u_i$ is comparable to $u_j$ for the \kl{prefix ordering}.
    Without loss of generality, because $\prefleq$ is well-founded,
    we can assume that for all $i < j$,
    $u_i \prefleq u_j$.

    Let $\aTransd = (Q, q_0, \delta, \lambda, F)$.
    Because $Q$ is a finite set,
    there exists $i < j$
    such that 
    $\delta(q_0, u_i) = \delta(q_0, u_j)$. Since 
    $u_i \prefleq u_j$, there exists $v \in \Sigma^*$
    such that $u_i v = u_j$.
    Let us now remark that for all $w \in \Sigma^*$,
    \begin{align*}
        f(u_j w) &= f(u_i v w) \\
                 &= \aTransd(q_0, u_i v w) \\
                 &= \aTransd(\delta(q_0, u_i v), w)
                   + \lambda(q_0, u_i v) (w) \\
                 &= \aTransd(\delta(q_0, u_i), w)
                   + \lambda(q_0, u_i v) (w) \\
                 &= f(u_i w) 
                   + \lambda(q_0, u_i v) (w)
    \end{align*}

    Because $\NPoly[k-1]$ is closed under shifts and finite sums,
    we conclude that
    $\app{f}{u_j} - \app{f}{u_i}$ belongs to $\NPoly[k-1]$,
    proving that $u_i \resleq{f}{k-1} u_j$.
\end{proof}

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item $f \in \NPoly[k]$;
        \item every $\prefleq$ monotone sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k-1}$.
        \item $f$ is \kl{computed} by a
            \kl{$k$-residual transducer};
    \end{enumerate}
\end{theorem}
\begin{proof}
    The first two items are known to be equivalent since 
    the phd of gaëtan.
    We have proven that transducers imply
    wqo, and that wqo implies $k$-residual transducer.
\end{proof}

\begin{corollary}
    In the \kl{commutative} case, one can effectively compute
    a \kl{$k$-residual transducer} of a function $f \in \NPoly$.
\end{corollary}

@@

\begin{theorem}
    The following are equivalent for $f \in \NPoly[k]$.
    \begin{itemize}
        \item The function $f \in \NSF[k]$,
        \item $f$ is \kl{ultimately polynomial}.
        \item Every \kl{$k$-residual transducer} of $f$ has no counters,
        \item $f$ is \kl{computed} by some
            \kl{$\NSF[k-1]$-transducer}.
    \end{itemize}
\end{theorem}
\begin{proof}
    The only implication to prove is the first implies the last.
    We know that the automata is counter free, so the languages of the states
    are star-free. Now, we just write 
    \begin{equation*}
        TODO
    \end{equation*}
    and we have finished the proof.
\end{proof}

>>>>

The goal of this section is to provide tools that do not require
\kl{commutative} \kl{$\Nat$-polyregular functions}. To that end, we introduce
the notion of \kl{$k$-residual transducer}, that generalizes the notion of
residual transducer introduced for \kl{$\Rel$-polyregular functions} in
\cite{LOPEZ23b}. We will first introduce the object, provide an algorithm (with
an oracle function) that computes said object, and finally, check that in
simple cases, the aperiodicity of a function can be read on its
\kl{$k$-residual transducer}. The main drawbacks of the generalization are the
following: one needs to decide whether a \kl{$\Rel$-polyregular function} is in
$\NPoly$, which is only known in the \kl{commutative} case, and the
characterization in terms of counter free also assumes \kl{commutativity}.
However, we believe that the introduction of this object is relevant.

@@

\begin{example}
    \label{non-canonical-transd:ex}
    Let $f \colon \set{a}^* \to \Nat$
    be defined as follows
    \begin{equation*}
        f(a^n) \defined
        \begin{cases}
            1 & \text{ if } n = 0 \\
            0 & \text{ if } n = 1 \\
            n & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}
    The function is computed by the two automata
    in \cref{non-canonical-transd:fig}.
\end{example}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.9\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);

            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
        \end{tikzpicture}
        \caption{This automaton has a \kl{counter}.}
    \end{subfigure}
    \begin{subfigure}[b]{0.9\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);

            \draw[->] (B) edge[loop right] node[midway, right]
                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
        \end{tikzpicture}
        \caption{This automaton is \kl{counter-free}.}
    \end{subfigure}
    \caption{Two \kl{$\NPoly[0]$-transducers}
    computing \cref{non-canonical-transd:ex}.}
    \label{non-canonical-transd:fig}
\end{figure}

\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.
An automaton is \reintro{counter-free} if it contains no \reintro{counters}.
This notion can be applied to \kl{$\mathcal{H}$-transducers} by
considering counters of their underlying automata.

@@

\AP Let us now introduce the notion of \intro{residual} of a function $f
\colon \Sigma^* \to \Rel$, by defining $\app{f}{u} \defined w \mapsto f(uw)$.
The collection of \reintro{residuals} of a function $f$ is denoted $\Res(f)$
and is defined as the set of $\app{f}{u}$ where $u$ ranges over words in
$\Sigma^*$.

\AP Let $k \in \Nat$. The residual transducer in the case of $\Rel$ output was
built using an equivalence relation defined as $f \equiv_k g$ if and only if $g
- f \in \ZPoly[k-1]$. However, in the case of $\Nat$ output, the relation
cannot be symmetric, and furthermore, we are going to order words rather than
residuals. Namely, given $k \in \Nat$, and $f \colon \Sigma^* \to \Nat$, we
define $u \resleq{f}{k} v$ if and only if $\app{f}{v} - \app{f}{u} \in
\NPoly[k-1]$.

\begin{fact}
    \label{good-residual-ordering:fact}
    Let $k \in \Nat$, and let $f \colon \Sigma^* \to \Nat$. Then,
    $(\resleq{f}{k})$ is a quasi-ordering, satisfying the following
    extra properties:
    \begin{enumerate}
        \item For all $u,v,w \in \Sigma^*$, $u \resleq{f}{k} v$
            implies $uw \resleq{f}{k} vw$,
        \item If $u \resleq{f}{k} v$ and $v \resleq{f}{k} u$,
            then $\app{f}{u} = \app{f}{v}$.
    \end{enumerate}
\end{fact}

There is one major issue arising when defining the notion of residual
transducer based on the quasi-ordering $\resleq{f}{k}$. The naïve approach is
to consider as states some representatives for the $\resleq{f}{k}$ minimal
elements of $\Sigma^*$, all computing the same function
(\cref{good-residual-ordering:fact}), and define transitions by letting
$\delta(u, a)$ be \emph{one} state $v$ such that $v \resleq{f}{k} ua$. This
naïve approach does not yield a canonical model.
For instance, the function
of \cref{non-canonical-transd:ex} can be computed in two ways if we only
consider $\resleq{f}{k}$ minimal elements, as witnessed in
\cref{non-canonical-transd:fig}. Even worse, the function $f$ belongs to
$\NSF[0]$, but one of the two presented transducers has a \kl{counter}.

This is why in \cref{residual-transducer:def}, we introduce the prefix ordering
$\prefleq$ to further constrain how the model computes.

\begin{definition}
    \label{residual-transducer:def}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
    is a \intro{k-residual transducer}
    of $f$ 
    when
    it is a \kl{$\NPoly[k-1]$-transducer}
    satisfying the following properties:
    \begin{enumerate}
        \item $\aTransd$ \kl{computes} $f$;
        \item $Q \subseteq \Sigma^*$ is a \kl{downwards closed}
            for $\prefleq$;
        \item $q_0 = \varepsilon$;
        \item every state $q \in Q$ is accessible from $q_0$;
        \item For all $u, a \in Q$,
            $\delta(u,a)$ is the $\prefleq$-maximal $v \in Q$
            such that $v \prefleq ua$, and $v \resleq{f}{k} ua$.
        \item For all $u,a \in Q$,
            $\lambda(u,a) = \app{f}{ua} - \app{f}{\delta(u,a)}$.
    \end{enumerate}
\end{definition}

Because of the restriction on prefixes, the \kl{$k$-residual transducer}
is actually unique when it exists.

\begin{fact}
    \label{unique-res-transducer:fact}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    Then $f$ has at most one \kl{$k$-residual transducer}.
\end{fact}
\begin{proof}
    Let $\aTransd_1$ and $\aTransd_2$ be two
    \kl{$k$-residual transducers} for $f$.
    The two initial states must be $\varepsilon$.
    Let us prove by induction on $u \in \Sigma^*$ that
    $\delta_1(\varepsilon, u) = \delta_2(\varepsilon, u)$
    and that $Q_1$ equals $Q_2$ over prefixes of $u$.
    This will prove that 
    $Q_1 = Q_2$, hence that $\aTransd_1 = \aTransd_2$.

    Let $u \in \Sigma^* \cap Q_1 \cap Q_2$ and $a \in \Sigma$, $v_1 \in Q_1$ be
    defined as $v \defined \delta_1(u,a)$, and $v_2 \defined \delta_2(u,a)$.
    Remark that by induction hypothesis, for all $v \prefle u$, $v \in Q_1 \cap
    Q_2$. If $\delta_1(u,a) = Q_1$, it means that for all $v \in Q_2$ such that
    $v \prefle ua$, we have $\neg( v \resleq{f}{k} ua )$. The only possible
    transition in $\aTransd_2$ is therefore $\delta_2(u,a) = ua$, and $ua \in
    Q_2$. Similarly, if $\delta_1(u,a) \prefleq u$, then $\delta_2(u,a) =
    \delta_1(u,a)$ by definition of $\delta_2$ as a maximum.
\end{proof}

We now introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
given a function $f$. Notice that this algorithm requires the ability to test
if a function belongs to $\NPoly[k]$, which is only known to be feasible for
\kl{commutative} polyregular functions. However, the termination of this
algorithm also proves the existence of the \kl{$k$-residual transducer}.

@@

        \eIf{$\exists v \in Q, v = \max_{\prefleq} \setof{w \in Q}{w \prefleq ua \wedge w \resleq{f}{k} ua}$}{
            $\delta(u, a) \defined v$;

@@

            $O \defined O \cup \setof{ uab }{b \in \Sigma}$;
        }
    }
    \For{$u \in Q$}{
        $F(u) \defined f(u)$;
    }
    \Return{$(Q, \varepsilon, \delta, \lambda, F)$};
    \caption{Computing a $k$-residual transducer given a function $f$.}
    \label{residual:algo}
\end{algorithm}

\begin{fact}
    \label{q-o-prefix-cool:fact}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    At each step of the \texttt{while loop}
    of \cref{residual:algo}, the sets
    $Q$ and $O$ are such that
    \begin{enumerate}
        \item $Q \cup O$ is a \kl{downwards closed} subset of 
            $\Sigma^*$ for $\prefleq$;
        \item elements in $O$ are pairwise incomparable
            for $\prefleq$, and are maximal
            for $\prefleq$ inside $Q \cup O$.
    \end{enumerate}
\end{fact}
\begin{proof}
    Let us write $Q_i$ and $O_i$ for the value of the variables
    $Q$ and $O$ at step $i$ of the \texttt{while loop}.
    We prove the desired property by induction on $i$.

@@

\begin{lemma}
    \label{correct-residual:lemma}
    If \cref{residual:algo} terminates on 
    an input $f \colon \Sigma^* \to \Nat$, $k \in \Nat$,
    then it computes the \kl{$k$-residual transducer} of $f$.
\end{lemma}
\begin{proof}
    Because of \cref{q-o-prefix-cool:fact},
    we already know that $q_0 = \varepsilon$,
    $Q$ is a \kl{downwards closed} subset of $\Sigma^*$
    for $\prefleq$, 
    that every state of $Q$ is accessible from $q_0$.
    Notice that at every step,
    $\lambda(u,a)$ is defined as
    $\app{f}{ua} - \app{f}{\delta(u,a)}$.
    Finally, since $Q \cup O$ is a \kl{downwards closed} subset of $\Sigma^*$
    at every step,
    we have that at step $i$,
    for all $ua \in O_i$,
    $\setof{w \in Q}{w \prefleq ua} = \setof{w \in Q_i}{ w \prefleq ua}$,
    which proves that the maximum considered in the algorithm
    is indeed computing correctly.
\end{proof}

\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
quasi-ordered set $(X, \leq)$ is \kl{good} whenever there exist $i < j$ such
that $u_i \leq u_j$. The set $X$ is a \kl{well-quasi-ordering} when
every infinite sequence is \kl{good}. A sequence is \kl{bad}
when it is not \kl{good}.

\begin{lemma}
    \label{wqo-implies-termination:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and $k \in \Nat$ be such that
    every infinite, $\prefleq$-increasing sequence is \kl{good}
    in $(\Sigma^*, \resleq{f}{k})$.
    Then, \cref{residual:algo} terminates on the input $(f,k)$.
\end{lemma}
\begin{proof}
    Assume towards a contradiction that
    \cref{residual:algo} does not terminate.
    Then, the \texttt{else} branch in the \texttt{while loop}
    must be taken infinitely often.
    This means that the set $Q$ of states grows arbitrarily large.

@@

    Let us prove that this infinite branch is a \kl{bad sequence} for the
    ordering $\resleq{f}{k}$.
    Let $j < p$, and assume by contradiction that $u_j \resleq{f}{k} u_p$. We
    know that $u_j \in Q_j$ and $u_p \in Q_p$. Then, at step $p-1$ of the
    algorithm, $u_j \in Q_{p-1}$, since $u_j \in Q_j \subseteq Q_{p-1}$.
    Because $u_j \prefleq u_p$ and $u_j \resleq{f}{k} u_p$,
    \cref{residual:algo} must take the \texttt{if} branch at step $p-1$. As a
    consequence, $u_p \not\in Q_{p}$, which is absurd.

    We have proven that the infinite branch is a \kl{bad sequence}
    for $\resleq{f}{k}$, which contradicts the assumption.
    Hence, \cref{residual:algo} must terminate.
\end{proof}

\begin{lemma}
    \label{n-poly-k-implies-wqo:lemma}
    Let $k \in \Nat$, and let $f \in \NPoly[k]$.
    Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
\end{lemma}
\begin{proof}
    Because $f \in \NPoly[k]$, there exists
    a tuple $\vec{x}$ of first order free variables,
    $\MSO$ formulas $\seqof{\psi_i(\vec{x})}{1 \leq i \leq n}$,
    and positive coefficients $\seqof{m_i}{1 \leq i \leq n}$,
    such that
    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$.

    Let $q$ be the maximal quantifier rank of formulas $\seqof{\psi_i}{1 \leq i
    \leq n}$. To a word $u \in \Sigma^*$, we associate the vector $\MSO^q(u)$
    that maps an $\MSO$-type with $\ell \leq |\vec{x}|$ free variables to the
    number of realizations of this type in $u$.

    Let $u, v \in \Sigma^*$ such that $\MSO^q(u) \leq \MSO^q(v)$, which means
    that every $\MSO$ type (of quantifier rank $q$ and with at most $n$ free
    variables) has at least as many realizations in $v$ than it has in $u$.
    Remark that by the compositionality of $\MSO$ over words (for instance, see
    the Feferman-Vaught theorem \cite{FEVAU59,MAKOW04}), for all $\MSO^q$ types
    $t(\vec{x})$, there are finitely many $\MSO^q$ types $t_l^j(\vec{y_i}),
    t_r^j(\vec{z_i})$ with $\vec{x} = \vec{y_i} \uplus \vec{z_i}$
    for $1 \leq j \leq N_0$, such that for every
    tuple $\vec{a}$ of elements in a word $uv$, $\MSO^q(\vec{a} / uv) =
    t(\vec{x})$ if and only if there exists $1 \leq j \leq N_0$,
    such that $\vec{a} = \vec{b} \uplus \vec{c}$,
    $\MSO^q(\vec{b} / u) =
    t_l^j(\vec{y_i})$, and $\MSO^q(\vec{c} / v) = t_r^j(\vec{z_i})$.
    We write $t = t_l \odot t_r$ to signify
    that $\MSO^q(\vec{bc} / uv) = t$
    if and only if $\MSO^q(\vec{b}/u) = t_l$
    and $\MSO^q(\vec{c}/v) = t_r$.

    As a consequence, if $\MSO^q(u) \leq \MSO^q(v)$, then 
    for all $w \in \Sigma^*$: 
    \begin{align*}
        & f(vw) - f(uw) \\
        &= 
        \sum_{i = 1}^n m_i
        \left[
            \vcount{\phi_i(\vec{x})} (vw) -
            \vcount{\phi_i(\vec{x})} (uw)
        \right] \\
        &= 
        \sum_{i = 1}^n
        m_i
            \sum_{\phi_i \in t(\vec{x})}
        \left[
            \vcount{t(\vec{x})}(vw)
            -
            \vcount{t(\vec{x})}(uw)
        \right] \\
        &= 
        \sum_{i = 1}^n
        m_i
        \sum_{1 \leq j \leq N_0}
        \sum_{\phi_i \in t_l^j(\vec{y}) \odot t_r^j(\vec{z})}
        \\
        &\quad\quad
        \underbrace{
        \left[
            \vcount{t_r^j(\vec{y})}(v)
            -
            \vcount{t_r^j(\vec{y})}(u)
        \right] 
    }_{ \in \Nat }
            \times 
            \vcount{t_l^j(\vec{z})}(w)
    \end{align*}

    We have proven that if $\MSO^q(u) \leq \MSO^q(v)$, then $u \resleq{f}{k}
    v$. Recall that $\Nat^p$ is a \kl{well-quasi-ordering} when endowed with
    the product ordering, and therefore that $\setof{\MSO^q(u)}{u \in
    \Sigma^*}$ is a \kl{well-quasi-ordering}.

    Let $\seqof{u_i}{i \in \Nat}$ be an infinite sequence of $\Sigma^*$.
    Without loss of generality, one can assume that for all $i \neq j$, $u_i
    \equiv_k u_j$, i.e., that the difference $\app{f}{u_i} - \app{f}{u_j}$
    belongs to $\ZPoly[k-1]$, since the latter has finite index. Thanks to the
    above remarks, there exists $i < j$ such that $\MSO^q(u_i) \leq
    \MSO^q(u_j)$. As a consequence, $g \defined \app{f}{u_j} - \app{f}{u_i} \in
    \NPoly$, and therefore $g \in \NPoly[k-1]$. We have proven that there
    exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
\end{proof}

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item $f \in \NPoly[k]$;
        \item $(\Sigma^*, \resleq{f}{k})$ is a
            \kl{well-quasi-ordering};
        \item every $\prefleq$-increasing sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k}$;
        \item $f$ is \kl{computed} by a
            \kl{$k$-residual transducer}.
    \end{enumerate}
    If $f$ is asumed to be \kl{commutative}, then all of the above
    properties are decidable, and the conversions effective.
\end{theorem}
\begin{proof}
    The first two items are known to be equivalent since 
    the phd of gaëtan.
    We have proven that transducers imply
    wqo, and that wqo implies $k$-residual transducer.
\end{proof}

\begin{lemma}
    \label{aperiodic-iff-residual:lem}
    Let $L$ be a regular language.
    The following are equivalent:
    \begin{enumerate}
        \item $L$ is aperiodic;
        \item The \kl{$0$-residual transducer} of $\ind{L}$
            is \kl{counter-free}.
    \end{enumerate}
\end{lemma}
\begin{proof}
    If the $0$-residual transducer is counter free,
    then $L$ is aperiodic because the minimal automata is also counter-free.

    % TODO
    \textbf{TODO Claim:}
    on every branch of the automata, the words have distinct residuals.

    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
    w_2) = t$.

    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
    strict prefix of $s_0$. It must exist, because otherwise the two are always
    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
    this contradicts the minimality of $u$.

    We have proven that $s_0$ and $s_1$ are comparable, hence they
    are equal.
    Finally, we have proven that $\delta(q, w_1) = s_0$,
    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.

    As a consequence $t$ and $q$ were comparable for the prefix relation,
    hence equal, and therefore $\delta(q, w) = q$.
\end{proof}

\begin{conjecture}
    Let $f \in \NPoly$. The following are equivalent,
    \begin{enumerate}
        \item $f \in \ZSF$,
        \item the \kl{$k$-residual transducer} of $f$
            is \kl{counter-free},
            and its labels belong to $\ZSF$.
    \end{enumerate}
\end{conjecture}
\begin{proof}
    If it has a counter free residual transducer recursively, then
    it clearly belongs to $\ZSF$.

    % TODO: mix orderings and trees and maximal elements :(

    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
    w_2) = t$.

    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
    strict prefix of $s_0$. It must exist, because otherwise the two are always
    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
    this contradicts the minimality of $u$.

    We have proven that $s_0$ and $s_1$ are comparable, hence they
    are equal.
    Finally, we have proven that $\delta(q, w_1) = s_0$,
    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.

    As a consequence $t$ and $q$ were comparable for the prefix relation,
    hence equal, and therefore $\delta(q, w) = q$.
\end{proof}

@@

\begin{theorem}
    The following are equivalent for $f \in \NPoly[k]$.
    \begin{itemize}
        \item The function $f \in \NSF[k]$,
        \item $f$ is \kl{ultimately polynomial}.
        \item Every \kl{$k$-residual transducer} of $f$ has no counters,
        \item $f$ is \kl{computed} by some
            \kl{$\NSF[k-1]$-transducer}.
    \end{itemize}
\end{theorem}
\begin{proof}
    The only implication to prove is the first implies the last.
    We know that the automaton is counter free, so the languages of the states
    are star-free. Now, we just write 
    \begin{equation*}
        TODO
    \end{equation*}
    and we have finished the proof.
\end{proof}
